A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-22 tex/body.tex
more

\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[scaled=.82]{beramono}
\usepackage[scaled=.88]{helvet}
\usepackage{booktabs}
\usepackage{bussproofs}
\usepackage{calc}
\usepackage{cite}
\usepackage{mathptmx}
%\usepackage{txfonts}
%\usepackage{mathrsfs}
%\usepackage{pifont}
%\usepackage{smallcap}
\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{subfigure}
\usepackage[usenames]{color}
%\usepackage{graphicx}
%\usepackage{newcent}
\usepackage{textcomp}
%\usepackage{tipa}
\usepackage{units}
\usepackage{url}
\usepackage[all]{xy}

@@

\noindent
of infinite sequences or streams over an element type $\tau$ is
allowed\begin{conf}, with no base case\end{conf}\begin{rep}; the
corresponding datatype would be rejected as non-well-founded
\cite{blanchette-et-al-2015-esop}\end{rep}.
%
Second, \emph{a uniqueness rule takes the place of the acyclicity rule of datatypes.}
Cyclic constraints such as
$\const{x} \teq \const{S}(\const{x})$ %, where $\const{C}$ is a constructor,
are unsatisfiable for datatypes but satisfiable for codatatypes.
However, the uniqueness principle states that two bisimilar values---i.e., two values
having the same possibly infinite expansion---must be equal; from $\const{x}
\teq \const{S}(\const{x})$ and
$\const{y} \teq \const{S}(\const{y})$, it deduces $\const{x} \teq \const{y}$.
The acyclicity and uniqueness rules are needed to ensure completeness
(model soundness) on ground problems. %and the absence of spurious models.
They cannot be expressed as finite axiomatizations, so they naturally belong in
a decision procedure.
%    * in particular, acyclicity and uniqueness are necessary for some proofs
%%      (and cannot be axiomatized finitely)
%    * and for model finding, without them we quickly get spurious models
%    * explain how finite model finding works
%(For the other (co)datatype properties---the injectivity, distinctness, and
%exhaustiveness of constructors and the selector laws---it is
%widely recognized that decision procedures can be more efficient than
%axiomatizations.)
%axiom.
%
Third, \emph{it must be possible to express cyclic }(\emph{regular}) \emph{values as closed terms and
to enumerate them.} This is necessary both for finite model finding
%modulo theories
and for theory combinations. The $\mu$-binder notation associates a name with
a subterm; it is used to represent cyclic values in the generated
models\begin{rep} and in the metatheory\end{rep}. For example,
the $\mu$-term $\const{SCons}(1,\: \MU s.\; \const{SCons}(0,\: \const{SCons}(9,\: s)))$
stands for the lasso-shaped sequence $1, 0, 9, 0, 9, 0, 9,
\ldots$\afterLdots.

@@

\ourparagraph{Phase 1: Computing the Bidirectional Closure \beforefig(\figurename~\ref{fig:cc-rules}).}
In conjunction with \rn{Refl}, \rn{Sym}, and \rn{Trans}, the \rn{Cong} rule computes the congruence (upward) closure,
whereas the \rn{Inject} and \rn{Clash} rules %together
compute the unification (downward) closure.
For unification, additional equalities are inferred based on the injectivity of constructors by \rn{Inject},
and failures to unify equated terms are recognized by \rn{Clash}.
The \rn{Conflict} rule recognizes when an equality and its negation both occur in $\Ec$, in which case $\Ec$ has no model.

@@

\ourparagraph{Phase 2: Applying Acyclicity and Uniqueness \beforefig(\figurename~\ref{fig:ab-rules}).}
%For presentation of the rules in this phase, we rely on a representation of terms in the $\mu$-notation.
The premises of the rules that make up the second phase refer to a mapping $\Val$
that associates a $\mu$-term with each equivalence class as its representative.
%Recall that $\mu$-bindings can be used for representing possibly cyclic terms and values.
%(Section~\ref{sec:introduction}).
% for example, $\MU x.\; \const{C}( \const{0}, x )$ represents the cyclic value $\const{C}( \const{0}, \const{C}( \const{0}, \ldots ))$.
The $\mu$-term $\Val[t\typ{\tau}]$ describes a class of $\tau$ values
that $t$ and other members of $t$'s equivalence class can take in models of $\Ec$.
%either in the case when $\tau$ is a codatatype or datatype type,
When $\tau$ is a datatype, % \in \Data$,
a cyclic $\mu$-term describes an infeasible class of values.
%$\tau \not\in \Codata$.

Formally, $\mu$-\emph{terms} are defined recursively as being either a variable $x$
or an applied constructor
$\MU x.\: \const{C}( \vec t\, )$ for some $\const{C} \in \Ctr$ and
$\mu$-terms $\vec t$ of the expected types.
The variable $x$ need not occur freely in the $\mu$-binder's body, in which case
the binder can be omitted.
In the \rn{Acyclic} rule,
$\FV( u )$ denotes the variables occurring freely in~$u$.
%%% But there are no constants of type $\tau \in \Nondata$ -- at most variables!
%
%For uniformity, we also consider $%\MU x.\;
%\const{c}$ to be a $\mu$-term if $\const{c}$ is a constant of type $\tau \in \Nondata$.
%
%%% I think this should be clear enough from the "Conventions" paragraph at the
%%% end of the introduction.
%%
% which we define recursively:
%if $\const{C} : \vec \tau \rightarrow \upsilon$
%and $\vec v$ are well-typed $\mu$-terms of type $\vec \tau$ under the assumption that $x$ has type $\upsilon$,
%then $\MU x.\; \const{C}( \vec v )$ is a well-typed term of type $\upsilon$.
%
A $\mu$-term is \emph{closed} if %and only if --- by convention, "and only if" is not necessary for *definitions* (but add it back if you disagree)
it contains no free variables. It is \emph{cyclic} if %and only if --- ditto
it contains a bound variable. The $\alpha$-\emph{equivalence} relation $t \aequiv u$
indicates that the $\mu$-terms $t$ and $u$
are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables\begin{rep}---e.g.,
%$\MU x.\; \const{C}( x ) \aequiv \MU y.\; \const{C}( y )$ and
$\MU x.\; \const{D}( y, x ) \aequiv \MU z.\; \const{D}( y, z )$,
but
$\MU x.\; \const{C}( x ) \not\aequiv \MU x.\; \const{D}( y, x ) \not\aequiv \MU x.\; \const{D}( z, x )
\not\aequiv \MU y.\; \const{D}( y, x )$\end{rep}.
Two $\mu$-terms can denote the same value despite being $\alpha$-disequivalent---e.g.,
$\MU x.\; \const{C}( x ) \not\aequiv \MU y.\; \const{C}( \const{C}( y ) )$.
%, or informally are \emph{observationally equivalent}.
%For convenience, we use $\mu$-terms to refer to (classes of) values for both codatatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

>>>>

\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[scaled=.82]{beramono}
\usepackage[scaled=.86]{helvet}
\usepackage{booktabs}
\usepackage{bussproofs}
\usepackage{calc}
\usepackage{cite}
\usepackage{mathptmx}
%\usepackage{txfonts}
%\usepackage{mathrsfs}
%\usepackage{pifont}
%\usepackage{smallcap}
\usepackage{mathpartir}
\usepackage{stmaryrd}
\usepackage{subfigure}
\usepackage[usenames]{color}
%\usepackage{graphicx}
%\usepackage{newcent}
\usepackage{textcomp}
%\usepackage{tipa}
\usepackage{units}
\usepackage{url}
\usepackage[all]{xy}

@@

\noindent
of infinite sequences or streams over an element type $\tau$ is
allowed\begin{conf}, with no base case\end{conf}\begin{rep}; the
corresponding datatype would be rejected as non-well-founded
\cite{blanchette-et-al-2015-esop}\end{rep}.
%
Second, \emph{a uniqueness rule takes the place of the acyclicity rule of datatypes.}
Cyclic constraints such as
$x \teq \const{S}(x)$ %, where $\const{C}$ is a constructor,
are unsatisfiable for datatypes but satisfiable for codatatypes.
However, the uniqueness principle states that two bisimilar values---i.e., two values
having the same possibly infinite expansion---must be equal; from $x
\teq \const{S}(x)$ and
$y \teq \const{S}(y)$, it deduces $x \teq y$.
The acyclicity and uniqueness rules are needed to ensure completeness
(model soundness) on ground problems. %and the absence of spurious models.
They cannot be expressed as finite axiomatizations, so they naturally belong in
a decision procedure.
%    * in particular, acyclicity and uniqueness are necessary for some proofs
%%      (and cannot be axiomatized finitely)
%    * and for model finding, without them we quickly get spurious models
%    * explain how finite model finding works
%(For the other (co)datatype properties---the injectivity, distinctness, and
%exhaustiveness of constructors and the selector laws---it is
%widely recognized that decision procedures can be more efficient than
%axiomatizations.)
%axiom.
%
Third, \emph{it must be possible to express cyclic }(\emph{regular}) \emph{values as closed terms and
to enumerate them.} This is necessary both for finite model finding
%modulo theories
and for theory combinations. The $\mu$-binder notation associates a name with
a subterm; it is used to represent cyclic values in the generated
models\begin{rep} and in the metatheory\end{rep}. For example,
the $\mu$-term $\const{SCons}(1,\: \MU s.\; \const{SCons}(0,\: \const{SCons}(9,\: s)))$
stands for the lasso-shaped sequence $1, 0, 9, 0, 9, 0, 9,
\ldots$\afterLdots.

@@

\ourparagraph{Phase 1: Computing the Bidirectional Closure \beforefig(\figurename~\ref{fig:cc-rules}).}
In conjunction with \rn{Refl}, \rn{Sym}, and \rn{Trans}, the \rn{Cong} rule computes the congruence (upward) closure,
whereas the \rn{Inject} and \rn{Clash} rules %together
compute the unification (downward) closure.
For unification, %additional
equalities are inferred based on the injectivity of constructors by \rn{Inject},
and failures to unify equated terms are recognized by \rn{Clash}.
The \rn{Conflict} rule recognizes when an equality and its negation both occur in $\Ec$, in which case $\Ec$ has no model.

@@

\ourparagraph{Phase 2: Applying Acyclicity and Uniqueness \beforefig(\figurename~\ref{fig:ab-rules}).}
%For presentation of the rules in this phase, we rely on a representation of terms in the $\mu$-notation.
The premises of the rules that make up the second phase refer to a mapping $\Val$
that associates with each equivalence class a $\mu$-term as its representative.
%Recall that $\mu$-bindings can be used for representing possibly cyclic terms and values.
%(Section~\ref{sec:introduction}).
% for example, $\MU x.\; \const{C}( \const{0}, x )$ represents the cyclic value $\const{C}( \const{0}, \const{C}( \const{0}, \ldots ))$.
The $\mu$-term $\Val[t\typ{\tau}]$ describes a class of $\tau$ values
that $t$ and other members of $t$'s equivalence class can take in models of $\Ec$.
%either in the case when $\tau$ is a codatatype or datatype type,
When $\tau$ is a datatype, % \in \Data$,
a cyclic $\mu$-term describes an infeasible class of values.
%$\tau \not\in \Codata$.

Formally, $\mu$-\emph{terms} are defined recursively as being either a variable $x$
or an applied constructor
$\MU x.\: \const{C}( \vec t\, )$ for some $\const{C} \in \Ctr$ and
$\mu$-terms $\vec t$ of the expected types.
The variable~$x$ need not occur freely in the $\mu$-binder's body, in which case
the binder can be omitted.
In the \rn{Acyclic} rule,
$\FV( u )$ denotes the variables occurring freely in~$u$.
%%% But there are no constants of type $\tau \in \Nondata$ -- at most variables!
%
%For uniformity, we also consider $%\MU x.\;
%\const{c}$ to be a $\mu$-term if $\const{c}$ is a constant of type $\tau \in \Nondata$.
%
%%% I think this should be clear enough from the "Conventions" paragraph at the
%%% end of the introduction.
%%
% which we define recursively:
%if $\const{C} : \vec \tau \rightarrow \upsilon$
%and $\vec v$ are well-typed $\mu$-terms of type $\vec \tau$ under the assumption that $x$ has type $\upsilon$,
%then $\MU x.\; \const{C}( \vec v )$ is a well-typed term of type $\upsilon$.
%
A $\mu$-term is \emph{closed} if %and only if --- by convention, "and only if" is not necessary for *definitions* (but add it back if you disagree)
it contains no free variables. It is \emph{cyclic} if %and only if --- ditto
it contains a bound variable. The $\alpha$-\emph{equivalence} relation $t \aequiv u$
indicates that the $\mu$-terms $t$ and $u$
are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables\begin{rep}---e.g.,
%$\MU x.\; \const{C}( x ) \aequiv \MU y.\; \const{C}( y )$ and
$\MU x.\; \const{D}( y, x ) \aequiv \MU z.\; \const{D}( y, z )$,
but
$\MU x.\; \const{C}( x ) \not\aequiv \MU x.\; \const{D}( y, x ) \not\aequiv \MU x.\; \const{D}( z, x )
\not\aequiv \MU y.\; \const{D}( y, x )$\end{rep}.
Two $\mu$-terms can denote the same value despite being $\alpha$-disequivalent---e.g.,
$\MU x.\; \const{C}( x ) \not\aequiv \MU y.\; \const{C}( \const{C}( y ) )$.
%, or informally are \emph{observationally equivalent}.
%For convenience, we use $\mu$-terms to refer to (classes of) values for both codatatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.
